NumPy---Array Computation for Python

### 2019-12-19 paper.tex
Fix typo
Add Governance section

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It underpins almost every library that does numerical
computation, including SciPy.  NumPy consists of the `ndarray`-object
along with utility functions to operate on such arrays.  Because of
its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy
array has become the {\it de facto} exchange format for array data in
Python.  That said, NumPy has now seen such widespread adoption that
not only the array object but also its {\it Application Programming
  Interface} (API) has become ubiquitous, and is imitated by, e.g.,
popular tensor computation libraries.

Over the past few years, with external financial support, NumPy has
seen renewed developer interest.  It is actively developed to meet the
needs of its community, including the addition of some fairly significant
changes as outlined in the Recent technical improvements section below.

@@

The Berkeley Institute for Data Science (BIDS) received two grants to support
the development of NumPy.
At the end of 2016, the Gordon and Betty Moore Foundation awarded them
\$645,020 for improving NumPy for better data science.
In 2017, the Alfred P. Sloan Foundation complemented the funding with an
additional \$659,359.
Stéfan van der Walt, a senior research data scientist at BIDS and a member of
the NumPy steering committee, is the PI of both grants and manages a team of
four programmers working on the project.

@@

NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
and OpenBLAS from the Chan Zuckerberg Initiative. 
NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
Ralf Gommers, the Director of Quansight Labs and a member of
the NumPy steering committee, is the PI.

@@

NumPy is now actively developed, with weekly online developer
meetings, a formalized NumPy enhancement proposal process, and
multiple active grants (but still almost no industry support).
% check almost no industry

@@

% \subsubsection*{Project goals / scope of library}

% https://numpy.org/neps/scope.html

% \subsubsection*{NEPS revived and roadmap}

@@

% https://numpy.org/neps/roadmap.html

NumPy, while having been wildly successful as a library and array exchange
project, has now also become a standard language for expressing array and
tensor computations.

@@

Broadly, the NumPy library consists of the following parts:
the NumPy array data structure; the so-called \emph{universal functions};
a set of library functions for manipulating arrays and doing scientific
computation; infrastructure libraries for unit tests and Python package
building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
The \code{ndarray} and the universal functions are generally considered
the core of the library.
In the following, we give a brief summary of these components of the
library.

\subsection*{Core.}

The \code{ndarray} data structure and the universal functions make up the core
of NumPy.

The \code{ndarray} is the data structure at the heart of NumPy.
The data structure stores regularly strided homogeneous data types
in a contiguous block memory, allowing for the efficient representation
of $n$-dimensional data.
More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

The \emph{universal functions}, or more concisely, \emph{ufuncs},
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two values and computes
$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
the ufunc will take care of looping over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-d array with length 3, and
\code{y} is a 2-d array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$.  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to complete the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.

@@

NumPy provides a large library of functions for array manipulation
and scientific computing, including functions for: creating, reshaping,
concatenating, and padding arrays; searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.

A suite of functions for computing the \emph{fast Fourier transform (FFT)}
and its inverse is provided.

NumPy's linear algebra library includes functions for: solving linear
systems of equations; computing various functions of a matrix, including
the determinant, the norm, the inverse, and the pseudo-inverse;
computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
and more.

The random number generator library in NumPy provides several alternative
\emph{bit stream generators} that provide the core function of generating
random integers.
A higher-level generator class that implements an assortment of
probability distributions is provided. It includes the beta, gamma
and Weibull distributions, the univariate and multivariate normal
distributions, and more.

A \emph{masked array} class, built on top of the \code{ndarray}, is
provided.  This class allows elements of an array to be masked,
indicating that the data is missing or unknown.

There is also a \code{matrix} class the implements 2-d arrays with
specialized behavior.  We mention it here for completeness, but the
use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
class have proven to be troublesome, and with the addition of the
infix \code{@} operator to Python for matrix multiplication, there was no
longer a need for a separate \code{matrix} class, since one can
use a 2-d array instead.

@@

NumPy provides utilities for writing unit tests and for building Python
packages.

The \code{testing} subpackage provides functions such as
\code{assert\_allclose(actual, desired)} that may be used in unit
test suites for code that uses NumPy arrays.

NumPy provides a subpackage includes functions and classes
that can be used by libraries that have NumPy as dependency,
to be used in the library's configuration code that installs the
library and that builds a package that can be used, for example,
on the PyPI web site.

The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

@@

In addition to the testing and building packages described above,
as the fundamental package for numerical and scientific computing
in Python, NumPy also plays a central role in building and
standardizing much of the community infrastructure.
The docstring standard mentioned in the history section is now
widely adopted.

% https://numpy.org/neps/nep-0029-deprecation_policy.html
In NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem
adopt a common ``time window-based'' policy for support of Python and
NumPy versions. This standard will simplify downstream project and
release planning.

@@

\noindent While this approach retains complete flexibility, the method
\code{np.random.default\_rng} can be used to instantize a \code{Generator} when
reproducibility is not needed.

@@

\subsection*{Governance}

% incl. CoC

\subsection*{Maintainers and contributors}

%  nr of core devs / commit rights) & contributors

NumPy is currently maintained a group of 23 contributors with commit rights
to the NumPy code base. Out of these 17 maintainers were active in 2019.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through github
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
With each release attracting more than a hundred new contributors.

@@

The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
PR merged into the NumPy master branch.
In this plot PR opened by current and old maintainers are indicated.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.
Over the course of its history, NumPy has attracted PR by 823 contributors.
However, its development still relies heavily on the relative small number
of stable maintainers, who share more than half of the contributions among
themselves.

% may be useful:
% - https://github.com/docathon/watchtower
% - https://github.com/machine-shop/oss_devkit/blob/master/plot_pr.py % (PR over time plot, there may be other better versions out there)

>>>>

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It underpins almost every library that does numerical
computation, including SciPy.  NumPy consists of the \code{ndarray}-object
along with utility functions to operate on it.  Because of
its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy
array is the {\it de facto} exchange format for array data in
Python.
NumPy, while having been wildly successful as a library and array exchange
project, has now also become a standard language for expressing array and
tensor computations.  
It has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous, and is
implemented by, for example, popular tensor computation libraries.

@@

The Berkeley Institute for Data Science (BIDS) received two grants to support
the development of NumPy.
At the end of 2016, the Gordon and Betty Moore Foundation awarded them
\$645,020 for improving NumPy for better data science.
In 2017, the Alfred P. Sloan Foundation complemented the funding with an
additional \$659,359.
Stéfan van der Walt, a senior research data scientist at BIDS and a member of
the NumPy steering council, is the PI of both grants and manages a team of
four programmers working on the project.

@@

NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
and OpenBLAS from the Chan Zuckerberg Initiative. 
NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
Ralf Gommers, the Director of Quansight Labs and a member of
the NumPy steering council, is the PI.

@@

@@

Over the past few years, with external financial support, NumPy has seen
renewed developer interest.
It is actively developed, with weekly online developer
meetings and a formalized NumPy enhancement proposal process.

@@

% https://numpy.org/neps/scope.html
% https://numpy.org/neps/roadmap.html

@@

The NumPy array data structure and universal functions, which we describe
below, make up the core of NumPy.
The NumPy library also includes
a set of library functions for manipulating arrays and doing scientific computation,
libraries for unit tests and Python package building,
and the program \code{f2py}\cite{peterson2009f2py} for wrapping Fortran code in Python.

\subsection*{Core}

The \code{ndarray} data structure stores regularly strided homogeneous data
types in a contiguous block memory, allowing for the efficient representation
of $n$-dimensional data.
A \emph{masked array} class, built on top of the \code{ndarray}, is
provided as well. This class allows elements of an array to be masked,
indicating that the data is missing or unknown.
More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

The \emph{universal functions} (or \emph{ufuncs})
are functions written in C that implement efficient looping over
NumPy arrays. An important feature of ufuncs is the built-in
implementation of \emph{broadcasting}.  For example, the function
\code{arctan2(x, y)} is a ufunc that accepts two values and computes
$\tan^{-1}(y/x)$.  When arrays are passed in as the arguments,
the ufunc will take care of looping over the dimensions of the inputs
in such a way that if, say, \code{x} is a 1-d array with length 3, and
\code{y} is a 2-d array with shape $2 \times 1$, the output will be
an array with shape $2 \times 3$.  The ufunc machinery takes care
of calling the function with all the appropriate combinations of
input array elements to complete the output array.
The elementary arithmetic operations of addition, multiplication, etc.,
are implemented as ufuncs, so broadcasting applies even in expressions
such as \code{x + y * z}.

@@

NumPy provides a large library of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
For historical reasons, NumPy also includes basic functionality for
linear algebra,
fast Fourier transforms and windowing,
pseudorandom number generators,
and polynomial fitting.
We will keep supporting these, but we will not expand beyond them.

%A suite of functions for computing the \emph{fast Fourier transform (FFT)}
%and its inverse is provided.

%NumPy's linear algebra library includes functions for: solving linear
%systems of equations; computing various functions of a matrix, including
%the determinant, the norm, the inverse, and the pseudo-inverse;
%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
%and more.
%
%The random number generator library in NumPy provides several alternative
%\emph{bit stream generators} that provide the core function of generating
%random integers.
%A higher-level generator class that implements an assortment of
%probability distributions is provided. It includes the beta, gamma
%and Weibull distributions, the univariate and multivariate normal
%distributions, and more.

%For historical reasons, we provide a \code{matrix} class the implements 2-d
%arrays with specialized behavior.
%But we strongly recommend you use a 2-d array and use the infix \code{@} operator
%for matrix multiplication.
%
%There is also a \code{matrix} class the implements 2-d arrays with
%specialized behavior.  We mention it here for completeness, but the
%use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
%class have proven to be troublesome, and with the addition of the
%infix \code{@} operator to Python for matrix multiplication, there was no
%longer a need for a separate \code{matrix} class, since one can
%use a 2-d array instead.

@@

Being at the base of the scientific Python stack, NumPy provides some basic
infrastructure for other packages in the scientific Python ecosystem.
The \code{numpy.testing} subpackage provides functions such as
\code{assert\_allclose(actual, desired)} that may be used in unit
test suites for code that uses NumPy arrays.
The \code{numpy.distutils} subpackage provides build support for C++, Fortran,
BLAS/LAPACK, and other relevant libraries for scientific computing.
The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

@@

In addition to the testing and building packages described above,
as the fundamental package for numerical and scientific computing
in Python, NumPy also plays a central role in building and
standardizing much of the community infrastructure.
The docstring standard mentioned in the history section is now
widely adopted.
We are also using the NEP system as a way to help coordinate
the larger scientific Python community.
% https://numpy.org/neps/nep-0029-deprecation_policy.html
For example, in NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem adopt a
common ``time window-based'' policy for support of Python and NumPy versions.
This standard will simplify downstream project and release planning.

@@

\noindent While this approach retains complete flexibility, the method
\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
reproducibility is not needed.

@@

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these 17 maintainers were active in 2019.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PR) each.\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through github
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
With each release attracting more than a hundred new contributors.

@@

The stacked plot in Figure~\ref{fig:prs-over-time} shows the number of
PR merged into the NumPy master branch.
In this plot PR opened by current and old maintainers are indicated.
Although the number of PRs being merged fluctuates,
the plot indicates an increased number of contributions over the past
years.
Over the course of its history, NumPy has attracted PR by 823 contributors.
However, its development still relies heavily on a relatively small number
of active maintainers, who share more than half of the contributions among
themselves.

% https://mail.python.org/pipermail/numpy-discussion/2015-October/073849.html
% https://github.com/numpy/numpy/pull/6352
NumPy adopted an official Governance Document on October~5,
2015\cite{NumPyProjectGovernance}.
Project decisions are usually made by consensus of interested contributors.
This means for many thing everyone is entrusted with veto power.
A Steering Council, currently composed of 12~members, facilitates this
process and oversees daily development of the project by contributing code
and reviewing contributions from the community.

% https://mail.python.org/pipermail/numpy-discussion/2018-July/078476.html
% https://github.com/numpy/numpy/pull/11865
NumPy's official Code of Conduct was approved on September~1, 2018\cite{NumPyCodeofConduct}.
In brief, we strive to:
\emph{be open};
\emph{be empathetic, welcoming, friendly, and patient};
\emph{be collaborative};
\emph{be inquisitive}; and
\emph{be careful in the words that we choose}.
The Code of Conduct also specifies how breaches can be reported and outlines
the process for responding to such reports.
